---
title: "EDA Donux truculus 2023"
subtitle: "Complementary analysis to advice and management report FEMP_AND_04"
author: "Mardones, M; Delgado, M"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
linkcolor: blue
output:
  html_document:
    keep_md: true
    toc: true
    toc_deep: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: cosmo
    fontsize: 0.9em
    linestretch: 1.7
    html-math-method: katex
    self-contained: true
    code-tools: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup1}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center',
                      dev = 'jpeg',
                      dpi = 300)
#XQuartz is a mess, put this in your onload to default to cairo instead
options(bitmapType = "cairo") 
# (https://github.com/tidyverse/ggplot2/issues/2655)
# Lo mapas se hacen mas rapido
```


```{r}
library(tidyverse)
library(ggridges)
library(readxl)
library(here)
library(lubridate)
library(readr)
library(ggthemes)
```

## Background


The following document and code intends to carry out a complementary
methodological Exploratory Data Analysis from survey data in coquina (*Donux truculus*), in this case, with a biological component like lengths structure.


## Set path

```{r}
data <- here("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data")
```

## Read Data Base

```{r}
size2017 <- read.csv("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Anterior a 2020/data_ieo_2017_def.csv", sep = ";")
size2018 <- read.csv("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Anterior a 2020/data_ieo_2018_def.csv", sep = ";")
size2019 <- read.csv("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Anterior a 2020/data_ieo_2019_def.csv", sep = ";")
size2020 <- read.csv("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Anterior a 2020/data_ieo_2020_def.csv", sep = ";")

size2021 <- read_excel("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Posterior 2020/Data_size_Coquina_2021.xlsx", sheet = "Coquina_donax")
size2022 <- read_excel("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Posterior 2020/Data_size_Coquina_2022.xlsx",  sheet = "Coquina_donax")
size2023 <- read_excel("~/IEO/DATA/Datos FEMP_AND_04/Coquina_Data/Posterior 2020/Data_size_Coquina_2023.xlsx",  sheet = "Coquina_Donax")
```

## Test dimension and names columns and diferences

```{r}
dim(size2017)
dim(size2018)
dim(size2019)
dim(size2020)
dim(size2021)

```


```{r}
names(size2017)
names(size2018)
names(size2019)
names(size2020)
names(size2021)

```

Same names. Could merge the DF
```{r}
size_17_20 <- rbind(size2017,
                    size2018,
                    size2019,
                    size2020)
# new dimension
dim(size_17_20)
names(size_17_20)
```

```{r}
glimpse(size_17_20)
```
## Change `Date` columns from `character`to `Date` format

```{r}
size_17_20$Date <- dmy(size_17_20$Date)
# separo los meses , dias y años
# Separar en columnas de día, mes y año
realdate <- as.Date(size_17_20$Date, format="%Y-%M-%D")

dfdate <- data.frame(Date=realdate)
ANO=as.numeric (format(realdate,"%Y"))
MES=as.numeric (format(realdate,"%m"))
DIA=as.numeric (format(realdate,"%d"))

size2<-cbind(dfdate,ANO,MES,DIA,size_17_20)
colnames(size2)
table(size2$ANO)

```

Now we test.
```{r}
table(size2$ANO)
```
## Viz

first glimpse. Red line is SL50 (10.8 cm to female (Delgado & Silva, 2016)
```{r}
nreg <- ggplot(size2 %>% 
                 select(-1), 
               aes(x=Size, 
                   y = as.factor(MES),
                  fill= as.factor(rastro)))+
  geom_density_ridges(stat = "binline", 
                      bins = 50, 
                      scale = 1.2,
                      alpha=0.7)+
  facet_wrap(.~ANO, ncol=4) +
  geom_vline(xintercept = 10.8, color = "red")+
  scale_fill_manual(values = c("#636363", "#2c7fb8", "#de2d26", "#756bb1", "#2ca25f"),
                       name="Rastro")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme_few()+
  xlab("Longitud (cm.)")+
  ylab("")+
  xlim(0,40)
#scale_x_discrete((limits = rev(levels(talla2021$ANO_ARR))))+
nreg
```
by beach


```{r}
nbeach <- ggplot(size2 %>% 
                 select(-1), 
               aes(x=Size, 
                   y = as.factor(MES),
                  fill= as.factor(Beach)))+
  geom_density_ridges(stat = "binline", 
                      bins = 50, 
                      scale = 1.2,
                      alpha=0.7)+
  facet_wrap(.~ANO, ncol=4) +
  geom_vline(xintercept = 10.8, color = "red")+
  scale_fill_viridis_d(option="F",
                       name="Beach")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme_few()+
  xlab("Longitud (cm.)")+
  ylab("")+
  xlim(0,40)
#scale_x_discrete((limits = rev(levels(talla2021$ANO_ARR))))+
nbeach
```

